Constraint Grammar as a SAT problem

### 2015-05-29 paper/nodalida2015.tex
more text for paper

\documentclass[11pt]{article}
\usepackage{nodalida2015}
\usepackage{times}
\usepackage{mathptmx}
%\usepackage{txfonts}
\usepackage{url}
\usepackage{latexsym}
\usepackage{cite}
\usepackage{authordate1-4}
\usepackage{multirow}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

% Uncomment for full paper
\section{Related work}
\label{sect:related}

Our work is inspired by \cite{lager98}, which presents constraint rules as a disjunctive logic program,
and \cite{lager_nivre01}, which reconstructs 4 different formalisms from a logical point of view.
\cite{lindberg_eineborg98ilp,asfrent14} use Inductive Logic Programming to learn CG rules from a tagged corpus.
\cite{lager01transformation} discusses ordering of the CG rules.

\section{CG as a SAT problem}
Let us demonstrate our approach with the following example in Spanish.

@@

%Without more context or additional rules we cannot fully disambiguate the passage,
%but unlike with standard CG, we get only legal combinations in one solution.

%If we add a word, \emph{la casa grande} `the big house' 
%and a rule that removes verb if followed by an adjective
%\texttt{REMOVE (v) IF (1 (adj)) ;}
%we will get a unique solution with \texttt{det n adj}.
%Standard CG will also remove the verb reading, 
%but the choice of the word \emph{la} doesn't contribute to

@@

\paragraph{Rules are unordered.}

We have experimented with different strategies:
\begin{itemize}
\item Maximise
% \footnote{Something about local maximum}
%The algorithm commits to a subset of clauses (e.g. \emph{all words must have more than 0 true analyses}), and tries to maximise the number of true variables without changing that subset. So it is a local maximum as regards to the given set of clauses. 
number of rules used---We discarded this strategy, because it was too strong: if a rule leads to a conflict in one case, it was not be applied anywhere.
\item Emulate sequential order---Enter clauses produced by each rule one by one, and assume the solver state reached so far is correct; if a conflict is introduced by new clauses, discard them and move on to next.
\item [\checkmark] Maximise number of rule applications---If there is a conflict, find the smallest number of rule applications to discard so that the conflict is solved.
\end{itemize}
The chosen heuristic produced the best results with examples in the order of tens or hundreds of rules.
Larger rule sets are yet to be tested.

@@

\paragraph{Rules disambiguate more.}
Considering our example phrase and rules, standard CG implementation
can only remove readings for the word which is tagged as \texttt{prn} or
\texttt{det}. The SAT-based implementation interprets the rules as
``determiner and verb together are illegal'', and is free to take action that concerns also the word in the condition (\texttt{1 n} or \texttt{1 v}).
This means that the rules are logically flipped: \texttt{REMOVE prn IF
  (1 n)} translates into the same logical formula as  \texttt{REMOVE n
  IF (-1 prn)}. A rule with more conditions can be written as many
rules, each condition taking its turn to be the target of removal or selection. \\

@@

However, the results using our approach should not be too different from the
existing ones---our preliminary evaluation against VISL CG-3 shows
promising results in this regard.

\section{Applications}

As described in Section~\ref{sec:eval}, our implementation is not
competetive with the state of the art. It is more fruitful to focus on
things where the SAT-based representation can make an impact.
One such thing is grammar analysis.

Traditionally, CG rules are applied deterministically and
sequentially; at the execution of rule, there is no way to
go back to earlier rules and undo them. Thus there is no way to know
what part of the input comes from the morphological analyser and what
from the CG.

@@

Table~\ref{table:time} shows the execution time compared to VISL CG-3. 
The number of rules in the grammar is more significant than the word
count; this holds also for the performance of VISL CG-3.
From the SAT solving side, maximisation is the most costly operation.
Parsing Don Quijote (384,155 words) with the rule set of 261 rules,
the function was called 147,253 times, and with 19 rules, 132,255 times.
%However, with both rule sets, half of the sentences in Don Quijote
%needed less than 6 calls of maximise, and 75 \% of the sentences
%needed less than 14.
The difference in the execution times suggests that there are other
reasons for the worse performance---this is to be expected,
SATCG is currently just a naive proof-of-concept implementation with no optimisations.

\begin{table}
  \centering
  \begin{tabular}{|l|c|c|c|}
     \hline
   \textbf{\# words} & \textbf{\# rules} &  \textbf{SATCG} & \textbf{VISL CG-3} \\ \hline
                384,155  & 19   &  23.1s   & 4.2s\\ %\hline
	        384,155  & 99   &  1m40.6s & 6.1s \\ %\hline
                384,155  & 261  &  2m36.7s & 10.7s \\ \hline
  \end{tabular}
  \caption{Execution times.}
  \label{table:time}
\end{table}

% \paragraph{Removing rules}
% \begin{table}
%   \centering
%   \begin{tabular}{|l|c|c|}
%      \hline
%     \textbf{\# rules} & \textbf{SATCG} & \textbf{VISL CG-3} \\
%     \hline 
% 	19 &  83.15 & 83.33 \\ \hline  % 19 out of 23 preselected by authors
% 	18 &  83.19 & 83.04 \\ \hline  % 18 out of 19
% 	17 &  83.201 & 83.03 \\ \hline   % 17 out of 18/19, no diff
% 	16 &  83.197 & 83.030 \\ \hline  % 16 out of 17/18, no diff
% 	15 &  83.188 & 83.022\\ \hline  % 15 out of 16/17. Tie with 14 for SATCG, chose to retain the rule that gave better results for VISL CG-3.
% 	14 &  83.180 & 83.005 \\ \hline  % Tie with 15 for SATCG.
% 	13 &  83.159 & 82.951 \\ \hline  % 13 out of 15 
% 	12 &  83.134 & 82.884\\ \hline  % 12 out of 14
% 	11 &  83.089 & 82.847\\ \hline  
% 	10 &  83.051 & 82.802\\ \hline  
% 	9  & 82.997 & 82.802 \\ \hline
% 	8  & 82.906 & 82.681 \\ \hline  % REMOVE:r_imp Imp IF (0 Subj) (-1* CnjSub) ; 
% 	7  & 82.743 & 82.594 \\ \hline  % SELECT:s_pro PrnIndep IF (1C VerbFin) ;
% 	6  & 82.556 & 82.398\\ \hline  %  REMOVE:r_v_v Verb IF (-1C Vblex) ;
% 	5  & 82.327 & 82.028 \\ \hline % SELECT:s_adv Adv IF (1 Cnj_Rel_End OR Prep OR Det)
% 	4  & 82.065 & 82.028 \\ \hline % REMOVE:r_vfin_prep VerbFin IF (0 Prep) (1 Inf) 
% 	3  & 81.387 & 81.233 \\ \hline % REMOVE:r_noun N    IF (-1 N OR Prn)
% 	2  & 80.23 & 79.97 \\ \hline % SELECT:s_pri_1 (pri p3) IF (0 (vblex imp p2 sg))
% 	1  & 76.69 & 76.62 \\ \hline 
% 	0  & 70.70 & 70.70 \\ \hline
%   \end{tabular}
%   \caption{Result of removing rules one by one, and how close to the gold standard they get.}
%   \label{table:results_remove}
% \end{table}
% In this experiment, we took 19 rules written by the authors, which give 83.15 \% correct results with SATCG and 83.33 \% with VISL CG-3.
% Then we removed rules one by one; after getting a score for $n$ rules, we took the combination of $n-1$ rules for which SATCG gives the best score, and we ran the rule set with VISL CG-3 for comparison. Table~\ref{table:results_remove} shows the percentages compared to the gold standard for both systems.

>>>>

\documentclass[11pt]{article}
\usepackage{nodalida2015}
\usepackage{times}
\usepackage{mathptmx}
\usepackage{fixltx2e}
%\usepackage{txfonts}
\usepackage{url}
\usepackage{latexsym}
\usepackage{cite}
\usepackage{authordate1-4}
\usepackage{multirow}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

\section{Related work}
\label{sect:related}

Our work is inspired by \cite{lager98}, which presents constraint rules as a disjunctive logic program,
and \cite{lager_nivre01}, which reconstructs 4 different formalisms from a logical point of view.
\cite{lindberg_eineborg98ilp,asfrent14} use Inductive Logic Programming to learn CG rules from a tagged corpus.
We explore different approaches to ordering. In the case of unordered
rules, it resembles Finite-State Intersection Grammar \cite{koskenniemi90}.
See also \cite{lager01transformation} for discussion of ordering of the CG rules.
\section{CG as a SAT problem}
Let us demonstrate our approach with the following example in Spanish.

@@

% Without more context or additional rules we cannot fully disambiguate the passage,
% but unlike with standard CG, we get only legal combinations in one solution.

@@

The traditional CG relies on that applying the rule set iteratively,
some rules fire the first time and make way for new rules to fire for
the second time.
For SAT-based CG, we present the rules as implications:
``if the \texttt{n} reading for \emph{casa} is true, then discard the
\texttt{prn} reading''. Each rule generates a list these implications,
and the SAT solver is left with the task of finding a model that will
satisfy them.

Let us continue with the earlier example. 
We can add a word to the sentence, \emph{la casa grande} `the big house',
and a rule that removes verb if followed by an adjective.
\begin{itemize}
\item [] \texttt{REMOVE v IF (1 adj) ;}
\end{itemize}
Then the additional clauses will make it impossible for \emph{casa} to
be a verb\footnote{If \texttt{adj} is the only reading for
\emph{grande}, it must be true, because of the restriction that at
least one analysis for each word is true. Then the implication
\texttt{grande<adj> => $\neg$casa<v>} has a true antecedent, thus
its consequent will hold.}, so we will get a unique solution with \texttt{det n adj}.
Standard CG will also remove the verb reading, 
but remember the beginning: if the \texttt{det} reading of \emph{la}
was removed first, then we are stuck with that choice.

% \todo{Unordered rules: also not applicable anymore to apply rules
% iteratively; there's no ``this rule doesn't fire now but will after
% applying X and Y'', it's all just implications ``this rule will fire
% if this is true'' and let the SAT solver find if those rules can apply
% peacefully to the same input.}

\paragraph{Rules can be unordered.}

As hinted by the previous property, SAT solver doesn't need a fixed
order of the rules.
Applying a rule to a sentence produces a number of clauses,
and those clauses are fed into the SAT solver.
However, in the unordered scheme, some information is lost: the
following rule sets would be treated identically, whereas in the
traditional CG, only the first would be considered as a bad order.

\begin{itemize}
\item [\texttt{1)}] \texttt{SELECT verb ;} \\
         \texttt{REMOVE verb IF (-1 det) ;} \\
         \texttt{REMOVE verb IF (-1 adj) ;} \\

\end{itemize}

\begin{itemize}
\item [\texttt{2)}] \texttt{REMOVE verb IF (-1 det) ;} \\
         \texttt{REMOVE verb IF (-1 adj) ;} \\
         \texttt{SELECT verb ;}
\end{itemize}

Without order, both of these rule sets will conflict, if applied to an
input that has sequence \texttt{det verb} or \texttt{adj verb}.
The SAT solver is fed clauses that tell to select a verb and remove a
verb, and it cannot build a model that satisfies all of those clauses.
To solve this problem, we create a variable for every clause
(\emph{instance of rule application}), and request a solution where maximally many of these variables are true.
If there is no conflict, then the maximal solution is one where all of
these variables are true; that is, all instances of rule application are kept.
In case of a conflict, SAT solver makes it possible to discard only
minimal amount of rule applications. For the example, assuming that
there are instances for both contexts for the REMOVE rules, then just
ignoring the SELECT rule will lead to maximal solution. 
This corresponds loosely to the common design pattern in
CGs, where there is a number of rules with the same target, ordered
such that more secure rules come first,
with a catch-all rule with no condition as the last resort, to be
applied if none of the previous has fired.

We can apply this for the whole grammar, or for one section at a time.

The design of unordered rules can lead to new ways of writing CGs,
potentially with less mental effort to the writer, but also less
transparent. By testing against gold standard, we see that the
unordered scheme not optimal for applying ready-made
CGs, where the order is a crucial detail.
We can also emulate order: enter clauses produced by each rule one by
one, and assume the solver state reached so far is correct; if a
conflict is introduced by new clauses, discard them and move on to
next.
This is also better for the performance, because the maximisation
problems are smaller.

%  that it is possible to write a small grammar, 
% which, when run with the unordered scheme, disambiguates more than the
% same grammar run with VISL CG-3. 
% However, when tested with grammars that are written for the original
% CG, SAT-CG loses with both strategies, but emulating order fares better.

The chosen heuristic produced the best results with examples in the order of tens or hundreds of rules.
Larger rule sets are yet to be tested.

@@

\paragraph{Rules disambiguate more.}
Considering our example phrase and rules, standard CG implementation
can only remove readings for the target word (\texttt{prn} or
\texttt{det}). The SAT-based implementation interprets the rules as
``determiner and verb together are illegal'', and is free to take action that concerns also the word in the condition (\texttt{1 n} or \texttt{1 v}).
This means that the rules are logically flipped: \texttt{REMOVE prn IF
  (1 n)} translates into the same logical formula as  \texttt{REMOVE n
  IF (-1 prn)}. A rule with more conditions corresponds to many
rules, each condition taking its turn to be the target of removal or selection. \\

@@

When conducting our tests, we have been able to construct a small grammar 
which, when run with the unordered scheme, disambiguates more than the
same grammar run with VISL CG-3. However, when tested with grammars
that are written for the original CG, SAT-CG loses with both ordering
strategies, but emulating order fares better. 
This suggests that our implementation should not compete with existing
state-of-the-art, but rather it has value as a way of relating the CG
formalism to wider context in the theory of computer science.
In Section~\ref{sec:apps} we discuss more about possible applications.

Table~\ref{table:time} shows the execution time compared to VISL CG-3. 
The number of rules in the grammar is more significant than the word
count; this holds also for the performance of VISL CG-3.
From the SAT solving side, maximisation is the most costly operation;
hence emulating ordering is faster, because maximisation problems are smaller.
Parsing Don Quijote (384,155 words) with the rule set of 261 rules,
the function was called 147,253 times, and with 19 rules, 132,255 times.
%However, with both rule sets, half of the sentences in Don Quijote
%needed less than 6 calls of maximise, and 75 \% of the sentences
%needed less than 14.
The difference in the execution times suggests that there are other
reasons for the worse performance---this is to be expected,
SATCG is currently just a naive proof-of-concept implementation with no optimisations.

\begin{table}
  \centering
  \begin{tabular}{|l|c|c|c|c|}
     \hline
   \textbf{\# words} & \textbf{\# rules} &  \textbf{SATCG\textsubscript{u}} & \textbf{SATCG\textsubscript{o}} & \textbf{VISL CG-3} \\ \hline
                384,155  & 19   & TODO &  23.1s   & 4.2s\\ %\hline
	        384,155  & 99   & TODO & 1m40.6s & 6.1s \\ %\hline
                384,155  & 261  & TODO & 2m36.7s & 10.7s \\ \hline
  \end{tabular}
  \caption{Execution times.}
  \label{table:time}
\end{table}

\section{Applications and Future Work}
\label{sec:apps}
As described in Section~\ref{sec:eval}, our implementation is not
competetive with the state of the art. It is more fruitful to focus on
things where the SAT-based representation can make an impact.
One such thing is grammar analysis.

Traditionally, CG rules are applied deterministically and
sequentially; at the execution of rule, there is no way to
go back to earlier rules and undo them. Thus there is no way to know
what part of the input comes from the morphological analyser and what
part from applying previous rules.

@@
